
\label{ch:tests}

Test cases for CISM include experiments with analytic solutions, standardized experiments
without analytic solutions but for which community benchmarks are available, and
some experiments specific to CISM which have been well-characterized by CISM developers.

Here we organize test cases based on the velocity solver that is more appropriate
for each test.  Note, however, that any velocity solver can be used with any test
if .config file settings are adjusted manually.  (However, in some cases the results
would be difficult to interpret.)

In general, each test directory includes a README file with some technical details 
about how to run the test.  Many tests have python scripts that are used to setup
the initial condition and, in many cases, actually execute the model.  Some tests
also have an additional python script for analyzing the CISM output in some way.

The user must manually provide each test with access to the CISM executable.
There are a few ways this can be done:

\begin{itemize}
  \item Softlink the executable into the directory, e.g.:

        \texttt{ln -s ../../builds/mac-gnu/cism\_driver/cism\_driver ./}

        This is the recommended procedure when doing development so that the test
        will always be using the most up-to-date version of the executable.

  \item Use the -e command line option to point to include an explicit path for the executable (for test case run scripts that support this option), e.g.:

        \texttt{./dome.py -e ../../builds/mac-gnu/cism\_driver/cism\_driver}

        This is useful for quickly trying a different version of CISM (e.g., comparing 
        serial and parallel executables).

  \item Add the directory that contains the CISM executable to your environment PATH.

  \item Copy the executable into the directory.  This is generally not the best practice,
        but may make sense it some situations.
\end{itemize}

For tests that use python scripts, the scripts generally have a number of useful
command line options that control their execution.  Typically, you can see details 
by using the --help (or -h) command line option, e.g.:

\texttt{./dome.py --help}



% =====================================
\input{\dir/shallow-ice-model-tests.tex}

% =====================================
\input{\dir/higher-order-model-tests.tex}

