\label{ch:runcism}

\section{Overview of Running CISM}

Assuming you successfully completed the Installation instructions in Chapter \ref{chp:install},
the executable for running the model, \texttt{cism\_driver}, can be found in your 
build directory in a subdirectory called \texttt{cism\_driver} 
(e.g., \texttt{./builds/mac-gnu/cism\_driver/cism\_driver}).

The build system creates the executable at this path but does not automatically
make it available to other locations on your system.  How you choose to do so depends 
on your situation.  See the introduction to Chapter \ref{chp:testcases} for 
an overview of how to make the executable available to other locations on your system
(e.g., symlinking, copying, or modifying your PATH environment variable).

Unlike previous versions of Glimmer, with CISM 2.0 this single \texttt{cism\_driver} 
executable is used for running the model in all configurations.
\texttt{cism\_driver} can be invoked with a single argument specifying 
a CISM .config file to run CISM as a standalone ice sheet model without Glint climate forcing,
or with two arguments (a CISM config file and a Glint config file) 
to run CISM with Glint climate forcing:
\begin{verbatim}
 Call cism_driver with either 1 or 2 arguments. Examples:
 cism_driver ice_sheet.config
 cism_driver ice_sheet.config climate.config
\end{verbatim}
The available options for the CISM configuration file and 
for the Glint climate interface configuration file are described in detail below.

To perform a parallel run with the parallel build of CISM, you must use the
MPI run command, which is typically \texttt{mpirun} or \texttt{mpiexec} but may 
vary among MPI versions and installations.  A standard syntax that is likely to
work on most installations is \newline
 \indent \texttt{mpirun -np N cism\_driver ice\_sheet.config \textit{climate.config}} \newline
where \texttt{N} is the number of processors you want to use, \texttt{ice\_sheet.config} is the name of the CISM
configuration file, and the optional argument \texttt{\textit{climate.config}} is the name 
of the climate configuration file.  For example: \newline
 \indent \texttt{mpirun -np 4 cism\_driver dome.config}\newline
would run the dome test case on four processors.

Finally, note that instructions for running CISM within the Community Earth System Model (CESM)
or another climate model are not described here.

When CISM runs, some basic information about its operation will be output to 
the screen (standard out).  More verbose information about the run will be written 
to a log file which is named \texttt{\textit{ice\_sheet.config}.log} where 
\texttt{\textit{ice\_sheet.config}} is the name of the .config file use to perform
the run.  (For example, if running the model with  \texttt{./cism\_driver dome.config},
the log file will be called  \texttt{dome.config.log}.)  The log file is an
important reference, especially for debugging runs that do not behave as expected.
For example, the log file includes a list of configuration options and parameter
values, which can be useful in diagnosing problems like typos in your .config file.
The log file also indicates what files were used for input and output and at which
times I/O occurred.  The log file may contain warnings about potentially
problematic configuration combinations or model behavior, such as the use of
configurations settings that are not scientifically validated, or a CFL violation
during advection.  In contrast, fatal errors will kill the model and the error
message will be written to both the screen and the log file.

Optionally, the log file also contains diagnostic information about the global
state of the ice sheet (e.g., the total ice area and volume, the maximum surface and basal
speeds, and the max and min temperatures), along with vertical profiles of speed
and temperature at a user-specified grid point.  This information is written at intervals
specified by the config file variable \texttt{dt\_diag}, for the diagnostic
point \texttt{(idiag,jdiag)}.

In addition to the log file, the model will create any netCDF output files requested
in the config file (see Section \ref{io-config} below for details).  
If the model dies for some reason midway through a simulation,
the netCDF files will still include output for the portion of the simulation that 
was completed.

\section{Overview of Configuration Files}

Running CISM is managed through configuration files (*.config) that enable 
desired model features and control input of initial conditions and forcing 
and output of model results.  This chapter summarizes the configuration options 
available for running CISM and is divided into sections on general Model Configuration, 
Input/Ouput Configuration, and optional Climate Forcing Configuration.

The format of CISM configuration files is taken from that used by the 
ConfigParser module in Python 2.x, which is similar to Windows \texttt{.ini} files 
and contains sections. Each section contains key/value pairs.

\begin{itemize}
\item[Comments:] Empty lines, or lines starting with a \texttt{\#}, \texttt{;} or \texttt{!} are ignored.  Comments can also be added on the same line as a key/value pair using these delimiters.
\item[Sections:] A new section starts with the section name enclosed by square brackets, \texttt{[ ]} and can be up to 50 characters long, e.g., \texttt{[grid]}.
\item[Key/Value Pairs:] Keys are separated from their associated values by \texttt{=} or \texttt{:}. The names can be up to 50 characters long. Values can be up to 400 characters long.
\end{itemize}

Sections and keys are case-sensitive and may contain white space. 
However, the configuration parser is very simple and thus the number of spaces 
within a key or section name also matters. Sensible defaults are used when 
a specific key is not found; defaults are shown in bold in the tables below.

Here is an example configuration file:
\begin{verbatim}
;a comment
[a section]
an_int  : 1
a_float = 2.0
a_char  = hey, this is rather cool
an_array = 10. 20. -10. 40. 100.

[another section]
! more comments
foo : bar
\end{verbatim}



